Gaps in stellar streams as a result of globular cluster fly-bys

### 2025-01-24 paper/main.tex
paola's pre sub final comments

\author{Salvatore Ferrone
       \inst{1,2}
         \and
       Marco Montuori\inst{2}
       \and
       Paola Di Matteo\inst{1}
       \and
       Alessandra Mastrobuono-Battisti \inst{3}
       \and 
       Misha Haywood \inst{1}
       \and
       Rodrigo Ibata \inst{4}
       \and 
       Paolo Bianchini \inst{4}
       \and
       Sergey Khoperskov \inst{5}
       \and
       Nicolas Leclerc \inst{1}
       \and 
       Clement Hottier \inst{1}
       \and 
       Eliot Stein  \inst{1}
       \and
       David Valls-Gabaud \inst{1}
       \and
       Owain N. Snaith \inst{6}
       }

\institute{
  Paris Observatory. Paris Sciences et Lettres\\
  \email{salvatore.ferrone@obspm.fr}
  \and
    Dipartimento di Fisica, Universit\`a di Roma ``La Sapienza'', Piazza Aldo Moro
  \and
    Dipartimento di Fisica e Astronomia. ``Galileo Gallilei'' Università di Padova, Vicolo dell'Osservatorio 3 Padova 35122, Italy.
    \and
      Universit\'e de Strasbourg, CNRS, Observatoire astronomique de Strasbourg, UMR 7550, F-67000 Strasbourg, France
    \and
      Leibniz-Institut für Astrophysik Potsdam (AIP), An der Sternwarte 16, 14482 Potsdam, Germany
    \and
      Exeter
    % \thanks{}
  }

@@

  These predictions occurred well before the first detections of Galactic globular cluster tidal tails \citep{1995AJ....109.2553G}. Interestingly, \citep{1995AJ....109.2553G}'s detections were made nearly contemporary  with the discovery of the Sagittarius stellar stream by \citet{1994Natur.370..194I} which is the closest example of a stream emerging from a dwarf satellite currently interacting with the Milky Way.

  Subsequent studies\footnote{For more subsequent observation detections of tidal debris, i.e. globular cluster stars beyond the tidal radius see the works of: \citet{1997A&A...320..776L, 2000A&A...356..127T, 2000A&A...359..907L, 2001AAS...19910906S, 2003AJ....126..815L}} extended Grillmair's findings to other globular cluster streams, but were often limited to the detections of stars still close to the cluster tidal radius, until the discovery made by \citet{2001ApJ...548L.165O,2002AAS...200.1001O, 2003AJ....126.2385O} of long and thin tails outside the Palomar~5 globular cluster. With a mass of $1.34\pm 0.24 \times 10^4 M_{\sun}$, Palomar~5 is one of the least massive globular clusters in the Galaxy. \citet{2003AJ....126.2385O} showed that its tails contain more mass than the cluster itself. The works of \citet{2006ApJ...641L..37G} and \citet{2015MNRAS.446.3297K} showed the tails to be extended for more than $20^\circ$ degrees in the sky. The discovery of its prominent tails stimulated a vigorous and successful search in the following years. New streams were discovered, mostly taking advantage of SDSS data, but also Pan-STARRS and ATLAS  \citep{2006ApJ...643L..17G, 2006ApJ...637L..29B, 2009ApJ...693.1118G, 2012ApJ...760L...6B, 2013ApJ...769L..23G, 2014ApJ...790L..10G, 2015ApJ...812L..26G, 2014MNRAS.443L..84B, 2016MNRAS.463.1759B, 2017ApJ...847..119G, 2014MNRAS.442L..85K}.

  \citet{2025NewAR.10001713B} provides a review of stellar stream astronomy, which has entered a new era since the publication of the data from the Gaia astrometric mission \citep{2016A&A...595A...1G}. Gaia's characterization of billions of stars in the Milky Way allows us to look for these structures coupling photometry, astrometry and, for the brightest stars, spectroscopy. The possibility given by Gaia to track stars with coherent movements, over the entirety of the sky, has led to the discovery of dozens of new streams. For example, \citet{2018MNRAS.477.4063M} developed the \texttt{streamfinder} algorithm and applied it across a series works of \citep{2018MNRAS.481.3442M,  2018ApJ...865...85I, 2019ApJ...872..152I} to discover the series of streams.%: Slidr, Sylgr, Ylgr, Fimbulthul, Svol, Gjoll, Leiptr, Phlegeton and Fjorm.  According to the most recent estimates provided by \citet{2020MNRAS.493.4978S} Fjorm in particular has an extension of more than 150 degrees on the sky which has a projected length more than five times greater than that of Palomar~5.

  The possibility to couple Gaia data with spectroscopic surveys, has extended the study of stellar streams beyond the quantification of their orbital properties, to a full chemical characterization. For example \citet{2019MNRAS.490.3508L} presents $S^5$, and is one example of a cross-catalog among other works \citep{2020AJ....160..181J, 2021ApJ...911..149L, 2022ApJ...928...30L, 2024MNRAS.529.2413U}. Currently, there are about a hundred stellar streams are currently known in our Galaxy. \citet{2023MNRAS.520.5225M} provides a compilation of their tracks on the sky. Interestingly, only about 20 streams are associated to known Galactic globular clusters.

@@

\section{Methods}
  The most accurate model of the Palomar~5 stream would involve full modeling of the internal dynamics of the cluster. This would mean computing $N$-Body interactions with a $\mathcal{O}(N^2)$ computation time, stellar evolution, supernovae, an initial mass distribution, treatment of binary star systems, etc. \citep[for such an example, see][]{2021NatAs...5..957G, 2016MNRAS.458.1450W}. Instead, we opt for solving the \textit{restricted-three body problem}, or as known as the \texttt{particle-test method} as we did for \citet{2023A&A...673A..44F}, which we describe here for completeness.

@@

    For the Galactic potential, we used the second model from \citet{2017A&A...598A..66P}, a superposition of a thin disk, thick disk, and dark matter halo, with masses and scale lengths provided in Table~1 of \citet{2023A&A...673A..44F}. This model is time-independent throughout our simulations. This model satisfies a series of observational constraints such as local solar stellar density, the Galactic rotation curve, etc, similarly to other Galactic models such as \texttt{MWpotential2014} from \citet{2015ApJS..216...29B} and \texttt{McMillian2017} from \citet{2017MNRAS.465...76M}. However, to balance data volume and computation time, we use only one Galactic potential, which should suffice. \citet{2021MNRAS.505.5978V} found that only a few outer globular clusters are strongly affected by different potential models. Generally, kinematic uncertainties are the dominant factor in orbital solution differences per cluster. Similarly, \citet{2024MNRAS.528.5189G} generated a globular cluster mass-loss catalog using seven different potential models and found their debris distributions to be rather model-independent, similarly to \citet{2023A&A...673A..44F}. While the clusters' positions may depend on the model, we assert that interaction rates and stream formation are unaffected by the choice of Galactic potential model.

@@

    We briefly compare our simulated gaps to the literature on Palomar~5. In the bottom panel of Fig.~\ref{fig:decomposition}, we compare the tracks of Palomar~5 that were compiled in Galstreams by \citet{2023MNRAS.520.5225M}. The mid-point positions of the streams do not have the same galacto-centric positions as Palomar~5 from Baumgardt's catalog and are thus offset when projected into tail coordinates. Moreover, since we sample the distances to Palomar~5, the galactocentric position within the Baumgardt catalog varies. To combat this, we position the mid-point of the Galstream tracks at the center of mass of the cluster, which allows us to compare the length of the observed tracks to our simulated streams. We use the three tracks Pal5-PW19, Pal5-S20, and Pal5-I21 from \citet{2019AJ....158..223P}, \citet{2020MNRAS.495.2222S}, and \citet{2021ApJ...914..123I}, respectively. For each track, we find the distance from the cluster in either direction and count the number of gaps that fall within this range. This is presented in Fig.~\ref{fig:GapsWithinSight}. In the maximum limit many gaps could appear, at a rate of about one per realization. However, at the shortest reported stream-length only a few gaps occur.
      \begin{figure}
        \centering
        \includegraphics[width=\linewidth]{GapWithinSight.png}
        \caption{The distribution of gaps as a function of absolute distance from the center of mass of Palomar 5's globular cluster. The number of gaps within the observable range is evaluated on an individual track basis, with track lengths determined by the minimum and maximum $x'$-coordinates from the bottom panel of Fig.~\ref{fig:decomposition}. The ranges for \citet{2019AJ....158..223P}, \citet{2020MNRAS.495.2222S}, and \citet{2021ApJ...914..123I} are shown for their respective dimensions, and the number of gaps within these ranges is indicated in the figure. Vertical bars mark the $x' > 0$ limit of the stream--we only show the right hand side instead of both for clarity, as most gaps are located on the right-hand side of the stream in any case. }
        \label{fig:GapsWithinSight}
      \end{figure}

@@

  The impact of globular clusters close interactions with stellar streams has also been the object of another recent paper, by \citet{2022ApJ...941..129D}, who concluded that the chance for GD-1 gaps, which have been reported in a number of works \citep[see, for example, ][]{2019ApJ...880...38B,2018MNRAS.477.1893D,2020AAS...23533607D} to be produced by globular clusters is very low. This result is not necessarily in contradiction with ours, since GD-1 has a pericenter which is almost twice that of Palomar~5 \citep[see, for example][]{2019MNRAS.486.2995M}. As we have discussed in Sect.~\ref{sect:geometry}, the number, $N$, of close encounters which lead to gap creation is anti-correlated with the distance $r$ to the Galactic center. If we naively use the same radial dependence of gaps from Palomar~5 for GD-1 by swapping a pericenter from 6~kpc to 16~kpc, we would reduced the number of gap-favorable impacts of more than a factor of 2. However, we note that \citet{2022ApJ...941..129D} pre-select the globular clusters which could have experienced a close encounter with the GD-1 stream, by selecting only clusters which pass at a distance less than 0.5 kpc from the stream, having a relative velocity smaller than  300 km/s. As shown in our Fig.~\ref{fig:impact_geometry_statistics}, bottom-left panel, in the case of Palomar~5 this choice would lead to exclude most of the encounters favorable to gap creation, which turn out to have relative velocities above 300 km/s. It would be interesting to repeat a similar study as the one made by \citet{2022ApJ...941..129D} for GD-1, this time imposing no selection on possible candidate clusters.\\

@@

  Future work should involve a systematic comparison of gap creation rates across various Galactic potentials and extend the scope to a wider range of streams and globular clusters. This would help identify regimes where baryonic and dark matter-induced substructures can be disentangled, shedding light on the elusive influence of dark matter sub-halos on stellar streams. Additionally, we can continue to add realistic physics to the Galactic environment such as the merger of the Sagittarius dwarf galaxy and the passages of the large and small Magellanic clouds.

@@

>>>>

\author{Salvatore Ferrone
       \inst{1,2}
         \and
       Marco Montuori\inst{2}
       \and
       Paola Di Matteo\inst{1}
       \and
       Alessandra Mastrobuono-Battisti \inst{3}
       \and 
       Rodrigo Ibata \inst{4}
       \and 
       Paolo Bianchini \inst{4}
       \and
       Sergey Khoperskov \inst{5}
       \and
       Nicolas Leclerc \inst{1}
       \and 
       Clement Hottier \inst{1}
       \and 
       Eliot Stein  \inst{6}
       \and
       Misha Haywood \inst{1}
       \and
       David Valls-Gabaud \inst{1}
       \and
       Owain N. Snaith \inst{7}
       }

\institute{
  LIRA, Observatoire de Paris, Université PSL, Sorbonne Université, Université Paris Cité, CY Cergy Paris Université, CNRS,75014 Paris, France\\
  \email{salvatore.ferrone@obspm.fr}
  \and
    Dipartimento di Fisica, Universit\`a di Roma ``La Sapienza'', Piazza Aldo Moro
  \and
    Dipartimento di Fisica e Astronomia. ``Galileo Gallilei'' Università di Padova, Vicolo dell'Osservatorio 3 Padova 35122, Italy.
  \and
    Universit\'e de Strasbourg, CNRS, Observatoire astronomique de Strasbourg, UMR 7550, F-67000 Strasbourg, France
  \and
    Leibniz-Institut für Astrophysik Potsdam (AIP), An der Sternwarte 16, 14482 Potsdam, Germany
  \and
    institute des sciences spatial
  \and
    Exeter
  }

@@

  These predictions occurred well before the first detections of Galactic globular cluster tidal tails \citep{1995AJ....109.2553G}. Interestingly, \citet{1995AJ....109.2553G}'s detections were made nearly contemporary  with the discovery of the Sagittarius stellar stream by \citet{1994Natur.370..194I} which is the closest example of a stream emerging from a dwarf satellite currently interacting with the Milky Way.

  Subsequent studies\footnote{For more subsequent observation detections of tidal debris, i.e. globular cluster stars beyond the tidal radius see the works of: \citet{1997A&A...320..776L, 2000A&A...356..127T, 2000A&A...359..907L, 2001AAS...19910906S, 2003AJ....126..815L}.} extended Grillmair's findings to other globular cluster streams, but were often limited to the detections of stars still close to the cluster tidal radius, until the discovery made by \citet{2001ApJ...548L.165O,2002AAS...200.1001O, 2003AJ....126.2385O} of long and thin tails outside the Palomar~5 globular cluster. With a mass of $1.34\pm 0.24 \times 10^4 M_{\sun}$ \citep{2019MNRAS.482.5138B}, Palomar~5 is one of the least massive globular clusters in the Galaxy. \citet{2003AJ....126.2385O} showed that its tails contain more mass than the cluster itself. The works of \citet{2006ApJ...641L..37G} and \citet{2015MNRAS.446.3297K} showed the tails to be extended for more than $20^\circ$ degrees in the sky. The discovery of its prominent tails stimulated a vigorous and successful search in the following years. New streams were discovered, mostly taking advantage of SDSS data, but also Pan-STARRS and ATLAS  \citep{2006ApJ...643L..17G, 2006ApJ...637L..29B, 2009ApJ...693.1118G, 2012ApJ...760L...6B, 2013ApJ...769L..23G, 2014ApJ...790L..10G, 2015ApJ...812L..26G, 2014MNRAS.443L..84B, 2016MNRAS.463.1759B, 2017ApJ...847..119G, 2014MNRAS.442L..85K}.

  \citet{2025NewAR.10001713B} provides a review of stellar stream astronomy, which has entered a new era since the publication of the data from the Gaia astrometric mission \citep{2016A&A...595A...1G}. Gaia's characterization of billions of stars in the Milky Way allows us to look for these structures coupling photometry, astrometry and, for the brightest stars, spectroscopy. The possibility given by Gaia to track stars with coherent movements, over the entirety of the sky, has led to the discovery of dozens of new streams. For example, \citet{2018MNRAS.477.4063M} developed the \texttt{streamfinder} algorithm and applied it across a series works \citep{2018MNRAS.481.3442M,  2018ApJ...865...85I, 2019ApJ...872..152I} to discover a multitude of streams.%: Slidr, Sylgr, Ylgr, Fimbulthul, Svol, Gjoll, Leiptr, Phlegeton and Fjorm.  According to the most recent estimates provided by \citet{2020MNRAS.493.4978S} Fjorm in particular has an extension of more than 150 degrees on the sky which has a projected length more than five times greater than that of Palomar~5.

  The possibility to couple Gaia data with spectroscopic surveys has extended the study of stellar streams beyond the quantification of their orbital properties, to a full chemical characterization. For example \citet{2019MNRAS.490.3508L} presents $S^5$, and is one example of a cross-catalog among other works \citep{2020AJ....160..181J, 2021ApJ...911..149L, 2022ApJ...928...30L, 2024MNRAS.529.2413U}. Currently, about a hundred stellar streams are known in our Galaxy. \citet{2023MNRAS.520.5225M} provides a compilation of their tracks on the sky. Interestingly, only about 20 streams are associated to known Galactic globular clusters.

@@

\section{Methods}
  The most accurate model of the Palomar~5 stream would involve full modeling of the internal dynamics of the cluster. This would mean computing $N$-body interactions with a $\mathcal{O}(N^2)$ computation time, stellar evolution, supernovae, an initial mass distribution, treatment of binary star systems, etc. \citep[for such an example, see][]{2021NatAs...5..957G, 2016MNRAS.458.1450W}. Instead, we opt for solving the \textit{restricted-three body problem}, or as known as the \texttt{particle-test method} as we did for \citet{2023A&A...673A..44F}, which we describe here for completeness.

@@

    For the Galactic potential, we used the second model from \citet{2017A&A...598A..66P}, a superposition of a thin disk, thick disk, and dark matter halo, with masses and scale lengths provided in Table~1 of \citet{2023A&A...673A..44F}. This model is time-independent throughout our simulations. This model satisfies a series of observational constraints such as local solar stellar density, the Galactic rotation curve, etc, similarly to other Galactic models such as \texttt{MWpotential2014} from \citet{2015ApJS..216...29B} and \texttt{McMillian2017} from \citet{2017MNRAS.465...76M}. However, to balance data volume and computation time, we use only one Galactic potential, which should suffice. \citet{2021MNRAS.505.5978V} found that only a few outer globular clusters are strongly affected by different potential models. Generally, kinematic uncertainties are the dominant factor in orbital solution differences per cluster. Similarly, \citet{2024MNRAS.528.5189G} generated a globular cluster mass-loss catalog using seven different potential models and found their debris distributions to be rather model-independent, similarly to \citet{2023A&A...673A..44F}. While the clusters' exact positions in time may depend on the model, we assert that interaction rates and stream formation are largely independent the choice of Galactic potential model.

@@

    We briefly compare our simulated gaps to the literature on Palomar~5. In the bottom panel of Fig.~\ref{fig:decomposition}, we compare the tracks of Palomar~5 that were compiled in \texttt{galstreams} by \citet{2023MNRAS.520.5225M}. The mid-point positions of the streams do not have the same galacto-centric positions as Palomar~5 from Baumgardt's catalog and are thus offset when projected into tail coordinates. Moreover, since we sample the distances to Palomar~5, the galactocentric position within the Baumgardt catalog varies. To combat this, we position the mid-point of the Galstream tracks at the center of mass of the cluster, which allows us to compare the length of the observed tracks to our simulated streams. We use the three tracks Pal5-PW19, Pal5-S20, and Pal5-I21 from \citet{2019AJ....158..223P}, \citet{2020MNRAS.495.2222S}, and \citet{2021ApJ...914..123I}, respectively. For each track, we find the distance from the cluster in either direction and count the number of gaps that fall within this range. This is presented in Fig.~\ref{fig:GapsWithinSight}. In the maximum limit many gaps could appear, at a rate of about one per realization. However, at the shortest reported stream-length only a few gaps occur.
      \begin{figure}
        \centering
        \includegraphics[width=\linewidth]{GapWithinSight.png}
        \caption{The distribution of gaps as a function of absolute distance from the center of mass of Palomar 5's globular cluster. The number of gaps within the observable range is evaluated on an individual track basis, with track lengths determined by the minimum and maximum $x'$-coordinates from the bottom panel of Fig.~\ref{fig:decomposition}. The ranges for \citet{2019AJ....158..223P}, \citet{2020MNRAS.495.2222S}, and \citet{2021ApJ...914..123I} are shown for their respective dimensions, and the number of gaps within these ranges is indicated in the figure. Vertical bars mark the $x' > 0$ limit of the stream--we only show the right hand side instead of both for clarity, as most gaps are located on the right-hand side of the stream in any case. }
        \label{fig:GapsWithinSight}
      \end{figure}

@@

  The impact of globular clusters close interactions with stellar streams has also been the object of another recent paper, by \citet{2022ApJ...941..129D}, who concluded that the chance for GD-1 gaps, which have been reported in a number of works \citep[see, for example, ][]{2019ApJ...880...38B,2018MNRAS.477.1893D,2020AAS...23533607D} to be produced by globular clusters is very low. This result is not necessarily in contradiction with ours, since GD-1 has a pericenter which is almost twice that of Palomar~5 \citep[see, for example][]{2019MNRAS.486.2995M}. As we have discussed in Sect.~\ref{sect:geometry}, the number, $N$, of close encounters which lead to gap creation is anti-correlated with the distance $r$ to the Galactic center. If we naively use the same radial dependence of gaps from Palomar~5 for GD-1 by swapping a pericenter from 6~kpc to 16~kpc, we would reduced the number of gap-favorable impacts of more than a factor of 2. Moreover, we note that \citet{2022ApJ...941..129D} pre-select the globular clusters which could have experienced a close encounter with the GD-1 stream, by selecting only clusters which pass at a distance less than 0.5 kpc from the stream, having a relative velocity smaller than  300 km/s. As shown in our Fig.~\ref{fig:impact_geometry_statistics}, bottom-left panel, in the case of Palomar~5 this choice would lead to exclude most of the encounters favorable to gap creation, which turn out to have relative velocities above 300 km/s. It would be interesting to repeat a similar study as the one made by \citet{2022ApJ...941..129D} for GD-1, this time imposing no selection on possible candidate clusters.\\

@@

  Future work should involve a systematic comparison of gap creation rates across various Galactic potentials and extend the scope to a wider range of streams and globular clusters. This would help identify regimes where baryonic and dark matter-induced substructures can be disentangled, shedding light on the elusive influence of dark matter sub-halos on stellar streams. Additionally, we can continue to add realistic physics to the Galactic environment such as the merger of the Sagittarius dwarf galaxy and the passages of the Large and Small Magellanic clouds.

@@

\begin{acknowledgements}
  This work has made use of the computational resources available at the Paris Observatory, as well as those obtained through the DARI grant A1020410154 (PI: P. Di Matteo), and of the Astropy, Numpy and Matplotlib libraries (Astropy Collaboration 2013, 2018; Harris et al. 2020; Hunter 2007). SF, PDM and MM would like to thank the Ecole Doctorale 127 for its financial support. MM thanks the CNR STM program, thanks to which the collaboration on this research topic began. 
\end{acknowledgements}
